---
output: github_document
bibliography: vignettes/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

[![Travis build status](https://travis-ci.org/venelin/PCMBase.svg?branch=master)](https://travis-ci.org/venelin/PCMBase)
[![Coverage status](https://codecov.io/gh/venelin/PCMBase/branch/master/graph/badge.svg)](https://codecov.io/github/venelin/PCMBase?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/PCMBase?color=blue)](https://cran.r-project.org/package=PCMBase)
[![Downloads](http://cranlogs.r-pkg.org/badges/PCMBase?color=blue)](https://cran.r-project.org/package=PCMBase)

# PCMBase : Simulation and likelihood calculation of phylogenetic comparative methods

Phylogenetic comparative methods represent models of continuous trait data associated with the tips of a phylogenetic tree. Examples of such models are Gaussian continuous time branching stochastic processes such as Brownian motion (BM) and Ornstein-Uhlenbeck (OU) processes, which regard the data at the tips of the tree as an observed (final) state of a Markov process starting from an initial state at the root and evolving along the branches of the tree. The PCMBase R package provides a general framework for manipulating such models. This framework consists of an application programming interface for specifying data and model parameters, and efficient algorithms for simulating trait evolution under a model and calculating the likelihood of model parameters for an assumed model and trait data. The package implements a growing collection of models, which currently includes BM, OU, BM/OU with jumps, two-speed OU as well as mixed Gaussian models, in which different types of the above models can be associated with different branches of the tree. Note that the PCMBase package does not implement model inference. Due to the enormous variety of models and possible model inference methods, this functionality is delegated to other packages that, taking advantage of PCMBase's fast likelihood calculation, can implement maximum likelihood (ML) or Bayesian inference methods. For example, the [PCMFit package](https://venelin.github.io/PCMFit/) provides heuristic-based ML fit of (mixed) Gaussian phylogenetic models [@Mitov:2019ci]. 

## Installation

### An optional but highly recommended dependency

The function `PCMTreePlot` in the package is implemented based on the R-package
ggtree, which is not on CRAN. It is highly recommended to install this package
in order to visualize trees with colored parts corresponding to 
defferent evolutionary regimes. If ggtree is not installed, the package will 
fail to run examples and generate the vignettes. At the time of writing this 
documentation, ggtree can be installed from bioconductor through the following 
code (if that does not work, check the 
[ggtree home page](https://guangchuangyu.github.io/software/ggtree/)):

```{r install-ggtree, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ggtree", version = "3.8")
```

### Installing PCMBase from CRAN

A stable but possibly old version of PCMBase is available on CRAN and can be 
installed with this command:
```{r eval=FALSE}
install.packages("PCMBase")
```

### Github
The newest but possibly less stable and tested version of the package can be 
installed using:
```{r eval=FALSE}
devtools::install_github("venelin/PCMBase")
```

## Resources

The user guides and technical reference for the library are available from the [PCMBase web-page](https://venelin.github.io/PCMBase/). 

* The [Getting started](https://venelin.github.io/PCMBase/articles/PCMBase.html)  guide
* The [PCMBase parameterizations](https://venelin.github.io/PCMBase/articles/PCMParam.html) guide
* The [Tracing the likelihood calculation of a Gaussian mode](https://venelin.github.io/PCMBase/articles/PCMTracePruning.html) guide
* The [Creating a Custom Model in the PCMBase Framework](https://venelin.github.io/PCMBase/articles/PCMCreateModel.html) guide

The research article "Fast likelihood evaluation for multivariate phylogenetic comparative methods: the PCMBase R package" provides a general overview of PCMBase. The article is currently undergoing peer review for a publication and is available as a preprint from [arxiv](https://arxiv.org/abs/1809.09014).

The PCMBase source code is located in the [PCMBase github repository](https://github.com/venelin/PCMBase). 

Feature requests, bugs, etc can be reported in the [PCMBase issues list](https://github.com/venelin/PCMBase/issues).

# Citing PCMBase

To give credit to the PCMBase package in a publication, please cite the following article:

Mitov, V., Bartoszek, K., Asimomitis, G., & Stadler, T. (2018, September 24). Fast likelihood evaluation for multivariate phylogenetic comparative methods: the PCMBase R package. arXiv.org. https://arxiv.org/abs/1809.09014.

# Used R-packages
```{r create-references, echo=FALSE, include=FALSE, eval=TRUE}
treeProcessing <- c("ape", "data.table")
maths <- c("expm", "mvtnorm")
plots <- c("ggtree", "ggplot2")
testing <- c("testthat", "covr")
docs <- c("roxygen2", "pkgdown")

packagesUsed <- c(treeProcessing, maths, plots, testing, docs)

printPackages <- function(packs) {
  res <- ""
  for(i in 1:length(packs)) {
    res <- paste0(res, paste0(packs[i], ' v', packageVersion(packs[i]), ' [@R-', packs[i], ']'))
    if(i < length(packs)) {
      res <- paste0(res, ', ')
    }
  }
  res
}

# Write bib information (this line is executed manually and the bib-file is edited manually after that)
knitr::write_bib(packagesUsed, file = "./vignettes/REFERENCES-R.bib")
```

The PCMBase R-package uses the following 3rd party R-packages:

* For tree processing in R: `r printPackages(treeProcessing)`;
* For algebraic manipulation: `r printPackages(maths)`;
* For plotting: `r printPackages(plots)`;
* For unit-testing: `r printPackages(testing)`;
* For documentation and web-site generation: `r printPackages(docs)`;

# Licence and copyright

Copyright 2016-2019 Venelin Mitov

Source code to PCMBase is made available under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version. PCMBase is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.

# References

