---
title: "Creating a Custom Model in the PCMBase Framework"
author: "Krzysztof Bartoszek"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Creating a Custom Model in the PCMBase Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
# Copyright 2016-2019 Venelin Mitov, Krzysztof Bartoszek
#
# This file is part of PCMBase.
#
# PCMBase is free software: you can redistribute it and/or modify
# it under the terms of the GNU General Public License as published by
# the Free Software Foundation, either version 3 of the License, or
# (at your option) any later version.
#
# PCMBase is distributed in the hope that it will be useful,
# but WITHOUT ANY WARRANTY; without even the implied warranty of
# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
# GNU General Public License for more details.
#
# You should have received a copy of the GNU General Public License
# along with PCMBase.  If not, see <http://www.gnu.org/licenses/>.
-->

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(abind)
library(PCMBase)
```
# Introduction
The PCMBase package is designed as a computational engine to calculate the likelihood for 
a user specified model evolving on a phylogenetic tree. The package comes
with some pre-implemented models, e.g. Brownian Motion or Ornstein-Uhlenbeck, both allowing
for multivariate correlated evolution.
However, the mathematical theory behind the method for calculating the likelihood 
(Mitov et. al. 2018) is not limited to these two models. In fact any 
stochastic process that belongs to the so-called GLInv family of models
is implementable in PCMBase. From a mathematical perspective one needs 
to be able to implement functions that calculate the compound
parameters $\vec{\omega}$, $\mathbf{\Phi}$ and $\mathbf{V}$. 
Furthermore, from the implementation side one has to be able to
handle the model defining parameters (default values, allowable
parametrizations, etc.). 

PCMBase's interface is based on the S3 object system (an excellent introduction 
by Hadley Wickham is available at http://adv-r.had.co.nz/S3.html).
Each implemented model will be an S3 object and below we will describe how a user
can implement a new model class by themselves.

# Brownian motion with drift
We will show how to create one's own model class with the example
of Brownian motion with drift. The model is a simple generalization
of the Brownian motion model by including a deterministic drift.
It is defined by the stochastic differential equation (SDE)
$$d\vec{X}(t) = \vec{h} dt + \Sigma_{x} d W(t),$$
where $W(t)$ is the standard Wiener process. Setting $\vec{h}=\vec{0}$
reduces the model to the Brownian motion process. 
The expectation of the process is $E[\vec{X}(t)] = \vec{X}(0)+t\vec{h}$
and the variance $Var[\vec{X}(t)] = t\Sigma_{x}\Sigma_{x}^{T}$.
Based on this we can calculate the the compound
parameters $\vec{\omega}$, $\mathbf{\Phi}$ and $\mathbf{V}$
defining the model in the GLInv family (Eq. 30, Mitov et. al. 2018)
$$
\vec{\omega} = \vec{h}t,
$$
$$
\mathbf{\Phi} = \mathbf{I},
$$
$$
\mathbf{V} = \Sigma_{x}\Sigma_{x}^{T}t.
$$ 
In the context of the PCM class setup we will actually have
$$
\mathbf{V} = \Sigma_{x}\Sigma_{x}^{T}t +  \Sigma_{e}\Sigma_{e}^{T}
$$ 
as these compound parameters are calculated at the end of branches.
The package allows for normally distributed $\vec{0}$ mean,
\Sigma_{e}\Sigma_{e}^{T} variance jumps at the end of a branch.

# A PCMBase class for Brownian motion with drift
We will now show how to implement the Brownian motion with
drift model in a class called "BM_drift that inherits
from the  "GaussianPCM" and "PCM" classes. It is easiest
if one takes an .R file from the PCMBase package that
already implements a model class and then modifies it accordingly.
We will work here with the BM.R file (that implements the BM model).
We need to redefine the generic functions of the "PCM" and "GaussianPCM"
classes to suit the Brownian motion with drift model.
We remind the reader that the function name is composed of two
parts separated by the dot. The first part is the generic
function, the second the name of the class in which it is implemented.
For example `PCMCond.BM_drift` is the `PCMCond` function's 
instance in the "BM_drift" class.

- `PCMParentClasses.BM_drift` : function returning the parental classes.
of "BM_drift":

```{r}
PCMParentClasses.BM_drift <- function(model) {
  c("GaussianPCM", "PCM")
}
```

- `PCMDescribe.BM_drift` : function returning a custom description of the model.

```{r}
PCMDescribe.BM_drift <- function(model, ...) {
  "Brownian motion model with drift"
}
```

- `PCMCond.BM_drift` : the key function when implementing a new model class.
This function returns a list with functions for calculating the $\vec{\omega}$,
$\mathbf{\Phi}$ and $\mathbf{V}$ compound parameters that define the 
conditional distribution of the daughter node given the value
of the trait at its parent node ($\vec{x}_{parent}$. Following Mitov et. al. (2018) we know 
that in the GLInv family a daughter node conditional on its parent is normally
distributed with expectation $\vec{\omega}+\mathbf{\Phi}\vec{x}_{parent}$ and
variance-covariance matrix $\mathbf{V}$. In particular the $\vec{\omega}$,
$\mathbf{\Phi}$ and $\mathbf{V}$ compound parameters can depend on time
(i.e. the length of the branch from the parental to the daughter node). In the
case of the BM with drift model the $\vec{\omega}=t\cdot \vec{h}$, 
$\mathbf{\Phi}$ is the identity matrix and $\mathbf{V}$ is the same as in the Brownian
motion model (which in itself is a limit of the OU model, hence the reuse of the 
PCMConcVOU function). 

```{r}
PCMCond.BM_drift <- function(
  tree, model, r = 1, metaI = PCMInfo(NULL, tree, model, verbose = verbose),
  verbose=FALSE) {

  
  Sigma_x <- if(is.Global(model$Sigma_x)){as.matrix(model$Sigma_x)}
	     else{as.matrix(model$Sigma_x[,, r])}
  Sigma <- Sigma_x %*% t(Sigma_x)
  if(!is.null(model$Sigmae_x)) {
    Sigmae_x <- if(is.Global(model$Sigmae_x)){as.matrix(model$Sigmae_x)}
		else{as.matrix(model$Sigmae_x[,,r])}
    Sigmae <- Sigmae_x %*% t(Sigmae_x)
  } else {
    Sigmae <- NULL
  }

  if(!is.null(model$h_drift)) { 
    h_drift <- if(is.Global(model$h_drift)) as.vector(model$h_drift) else model$h_drift[, r]
  }else{
    h_drift <- rep(0,nrow(Sigma_x))
  }

  V <- PCMCondVOU(matrix(0, nrow(Sigma), ncol(Sigma)), Sigma, Sigmae)
  omega <- function(t, edgeIndex, metaI) {
    t*h_drift
  }
  Phi <- function(t, edgeIndex, metaI, e_Ht = NULL) {
    diag(nrow(Sigma))
  }
  list(omega = omega, Phi = Phi, V = V)
}
```

- `PCMDescribeParameters.BM_drift` : a function that returns a list with a custom description
of each model parameter.

```{r}
PCMDescribeParameters.BM_drift <- function(model, ...) {
  list(
    X0 = "trait values at the root",
    h_drift = "drift vector modifying the expectation",
    Sigma_x = "Cholesky factor of the unit-time variance rate",
    Sigmae_x = "Cholesky factor of the non-heritable variance 
		or the variance of the measurement error")
}
```

- `PCMListParameterizations.BM_drift` : a function that returns all possible 
parametrizations for the implemented model class. These parametrizations 
correspond to how each parameter defining the model can be parametrized.
For a detailed description of how matrices can be parametrized please see
Bartoszek (2011), Bartoszek et. al. (2012) or the matrixparametrizations.R file
in the mvSLOUCH or PCMBase R packages. Probably from the perspective of just
calculating the likelihood given some parameters this function does not seem
that useful. However, one should not forget that PCMBase is designed
to be a computational engine providing the likelihood that will be optimized
over some other code. 
Here, $\vec{X}(0)$ and $\vec{h}$ are vectors (i.e. "VectorParameter"). 
The "_AllEqual" parametrization means all the entries of the 
vector should be equal. "_Global" means that it is the same for all regimes
(notice that $\vec{X}(0)$ is the value at the root, so it has to be common for all), 
"_Omitted" means not present (for $\vec{h}$ this means that the model will
correspond to a Brownian motion with $0$ drift). Finally "_Fixed" means that
the parameter is "known", i.e. it is not to be optimized over.
Matrix parametrizations are more involved. For example one will not optimize
over a covariance matrix but e.g. over its Cholesky decomposition.
Here we just have the $\Sigma_{x}$ and $\Sigma_{e}$ matrices
(however see the "OU" class for more involved cases with the $H$ 
matrix). Here, these two matrices can be either upper triangular,
diagonal or diagonal with a constant value on the diagonal. Both
of them enter the likelihood (through $\mathbf{V}$) as a product
of themselves and their transposition i.e. $SS^{T}$ so decomposition
into a triangular matrix (with non-negative diagonal) suffices for
unique identification of the matrix. Notice that if $S$ is of 
dimension $k\times k$, then $SS^{T}$ has $k^{2}$ potentially non-zero
elements while its triangular decomposition only $k(k+1)/2$
A detailed description of the different possible parametrizations is
given in the help file for PCMParam.

```{r}
PCMListParameterizations.BM_drift <- function(model, ...) {
  list(
    X0 = list(
      c("VectorParameter", "_Global"),
      c("VectorParameter", "_Fixed", "_Global"),
      c("VectorParameter", "_AllEqual", "_Global"),
      c("VectorParameter", "_Omitted")),
    h_drift = list(     
      c("VectorParameter"),
      c("VectorParameter", "_Fixed"),
      c("VectorParameter", "_AllEqual"),
       c("VectorParameter", "_Omitted")),
       
    Sigma_x = list(
      c("MatrixParameter", "_UpperTriangularWithDiagonal", "_WithNonNegativeDiagonal"),
      c("MatrixParameter", "_Diagonal", "_WithNonNegativeDiagonal"),
      c("MatrixParameter", "_ScalarDiagonal", "_WithNonNegativeDiagonal")),

    Sigmae_x = list(
      c("MatrixParameter", "_UpperTriangularWithDiagonal", "_WithNonNegativeDiagonal"),
      c("MatrixParameter", "_Diagonal", "_WithNonNegativeDiagonal"),
      c("MatrixParameter", "_ScalarDiagonal", "_WithNonNegativeDiagonal"),
      c("MatrixParameter", "_UpperTriangularWithDiagonal", "_WithNonNegativeDiagonal", "_Global"),
      c("MatrixParameter", "_Diagonal", "_WithNonNegativeDiagonal", "_Global"),
      c("MatrixParameter", "_ScalarDiagonal", "_WithNonNegativeDiagonal", "_Global"),
      c("MatrixParameter", "_Omitted"))
  )
}
```

- `PCMListDefaultParameterizations.BM_drift` : this function is optional to define but can
be useful if only a subset of the parametrizations defined in the PCMListParametrizations
function will actually be used in practise. 

```{r}
PCMListDefaultParameterizations.BM_drift <- function(model, ...) {
  list(
    X0 = list(
      c("VectorParameter", "_Global"),
      c("VectorParameter", "_Omitted")
    ),
    h_drift = list(     
      c("VectorParameter")),
       
    Sigma_x = list(
        c("MatrixParameter", "_UpperTriangularWithDiagonal", "_WithNonNegativeDiagonal"),
        c("MatrixParameter", "_Diagonal", "_WithNonNegativeDiagonal"),
        c("MatrixParameter", "_ScalarDiagonal", "_WithNonNegativeDiagonal")
      ),

    Sigmae_x = list(
      c("MatrixParameter", "_Omitted"))
  )
}
```

- `PCMSpecify.BM_drift` : generate default model parameters. Notice that here
we obtain a singular model with 0 mean and 0 variance.

```{r}
PCMSpecify.BM_drift <- function(model, ...) {
  spec <- list(
    X0 = structure(0.0, class = c('VectorParameter', '_Global'),
                   description = 'trait values at the root'),
    h_drift = structure(0.0, class = c('VectorParameter'),
                   description = 'drift vector modifying the expectation'),
    Sigma_x = structure(0.0, class = c('MatrixParameter', '_UpperTriangularWithDiagonal', 
					'_WithNonNegativeDiagonal'),
                        description = 'Cholesky factor of the unit-time variance rate'),
    Sigmae_x = structure(0.0, class = c('MatrixParameter', '_UpperTriangularWithDiagonal',
					'_WithNonNegativeDiagonal'),
                         description = 'Cholesky factor of the non-heritable variance 
                        		or the variance of the measurement error'))
  attributes(spec) <- attributes(model)
  if(is.null(names(spec))) names(spec) <- c('X0', 'h_drift', 'Sigma_x', 'Sigmae_x')
  if(any(sapply(spec, is.Transformable))) class(spec) <- c(class(spec), '_Transformable')
  spec
}
```

# Example run
Now that we have defined all the code necessary for the class let us demonstrate it.
After running all the above code defining the "BM_drift" class we create a model instance.
We do this for a two regimes model. 

```{r}
X0 <- c(5, 2, 1) ## root state
  
## in regime a traits evolve independently
a.Sigma_x <- rbind(c(1.6, 0.0, 0.0),c(0.0, 2.4, 0.0),c(0.0, 0.0, 2.0))
## no jumps at the end of a branch
a.Sigmae_x <- rbind(c(0.0, 0.0, 0.0),c(0.0, 0.0, 0.0),c(0.0, 0.0, 0.0))
a.h_drift<-c(4, 5, 6)

## in regime b evolution is correlated
b.Sigma_x <- rbind(c(1.6, 0.3, 0.3), c(0.0, 0.3, 0.4),c(0.0, 0.0, 2.0))
## no jumps at the end of a branch
b.Sigmae_x <- rbind(c(0.0, 0.0, 0.0),c(0.0, 0.0, 0.0),c(0.0, 0.0, 0.0))
b.h_drift<-c(1, 2, 3)

Sigma_x <- abind(a.Sigma_x, b.Sigma_x, along=3, new.names=list(x=NULL,y=NULL,regime=c('a','b')))
Sigmae_x <- abind(a.Sigmae_x,b.Sigmae_x,along=3,new.names=list(x=NULL,y=NULL,regime=c('a','b')))
h_drift <- abind(a.h_drift, b.h_drift, along=2, new.names=list(xy=NULL, regime=c('a','b')))

PCMBase_model_BM_drift <- PCM("BM_drift", k = 3, regimes = c("a", "b"),
params = list(X0 = X0,h_drift = h_drift[,,drop=FALSE],
Sigma_x = Sigma_x[,,,drop=FALSE],Sigmae_x = Sigmae_x[,,,drop=FALSE]))
```


Now we simulate a random phylogeny (here pure birth tree with unit speciation rate 
conditioned on $30$ tips). Furthermore, we randomly paint the two regimes, "a" and
"b" on the tips of the tree. Then, we use mvSLOUCH's (modified from slouch, Hansen et. al. 2008,
to work with phylo format trees) Fitch algorithm to paint the regimes on the internal 
branches of the tree. It is worth pointing out that the Fitch algorithm is only
one option and the user should explore other possibilities depending on the
problem.

```{r}
set.seed(12345, kind = "Mersenne-Twister", normal.kind = "Inversion") ## set a random seed
N<-30 ## number or tip species
phyltree<-ape::rcoal(n=N)

v_current_regimes<-sample(c("a","b"),30,replace=TRUE)
v_branch_regimes<-mvSLOUCH::fitch.mvsl(phyltree,v_current_regimes)$branch_regimes
v_ambiguous<-which(v_branch_regimes=="ambiguous")
if (length(v_ambiguous)>0){
    v_branch_regimes[v_ambiguous]<-names(which.max((table(v_branch_regimes[-v_ambiguous]))))
}
## some branches may have "ambiguous" assigned to them due to how the implementation of 
## Fitch's algorithm works. These branches need to have the regimes set manually. 
## We do this according to the maximum rule but in a real analysis this would have to be 
## done based on a careful analysis of the biological system under study.
phyltree<-PCMTreeSetRegimesForEdges(phyltree, v_branch_regimes,FALSE) 

```

We simulate the traits using PCMBase's functionality.

```{r}
mData<-PCMSim(phyltree, PCMBase_model_BM_drift, X0)[,1:N] ## we only want the tip data
## NOTE that observations from different species are in the columns NOT in the rows as 
## in other software
```

Finally we calculate the likelihood under the BM with drift model.

```{r}
log_lik<- PCMLik(mData, phyltree, PCMBase_model_BM_drift)
print(log_lik[1]) ## we just want to print the log-likelihood without the attributes
```

If PCMBase is used as a computational engine for some inference package, then the 
above code can be one way of obtaining the likelihood under a "GaussianPCM" model object. 
However, in such a setup it is more recommended to use the mechanism of creating a likelihood
function for a particular dataset through PCMCreateLikelihood. This will speed up the calculations
as it avoids traversing the tree every time the likelihood value is required. One just
needs to update the parameters to obtain a new likelihood value.

```{r}
## create an vector of appropriate length to store the vectorized model parameters
v_param <- double(PCMParamCount(PCMBase_model_BM_drift))

# load the current model parameters into param
PCMParamLoadOrStore(PCMBase_model_BM_drift, v_param, offset=0, load=FALSE)

print(v_param)

## now create a likelihood function for the particular model and observed data
likFun <- PCMCreateLikelihood(mData, phyltree, PCMBase_model_BM_drift)

log_lik_from_likFun<-likFun(v_param)
print(log_lik_from_likFun[1])
print(log_lik_from_likFun[1]==log_lik[1])

# modify slightly the model parameters
v_param_2 <- jitter(v_param)

print(v_param_2)

# set the new parameter vector
PCMBase_model_BM_drift_2<-PCMBase_model_BM_drift
PCMParamLoadOrStore(PCMBase_model_BM_drift_2, v_param_2, offset = 0, load=TRUE)

print(PCMBase_model_BM_drift_2)
log_lik_from_likFun_2<-likFun(v_param_2)
print(log_lik_from_likFun_2[1])

```

