---
title: "The PCMBase Parametrization API"
author: "Venelin Mitov"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{The PCMBase Parametrization API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
# Copyright 2016 Venelin Mitov
#
# This file is part of PCMBase.
#
# PCMBase is free software: you can redistribute it and/or modify
# it under the terms of the GNU General Public License as published by
# the Free Software Foundation, either version 3 of the License, or
# (at your option) any later version.
#
# PCMBase is distributed in the hope that it will be useful,
# but WITHOUT ANY WARRANTY; without even the implied warranty of
# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
# GNU General Public License for more details.
#
# You should have received a copy of the GNU General Public License
# along with PCMBase.  If not, see <http://www.gnu.org/licenses/>.
-->

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(abind)
library(PCMBase)
```

# Introduction
Model parameters are the key object of interest in every phylogenetic comparative method. PCMBase provides a powerful interface for specifying and manipulating model parameters. This interface is based on the S3 object system (see http://adv-r.had.co.nz/S3.html for an excellent introduction by Hadley Wickham).

In PCMBase, every model is an object of an S3 class, such as "OU", inheriting from the base S3-class "PCM". A PCM object represents a named list. Each element of that list can be one of the following:

- a global parameter shared by all regimes in the model. For example, this would
be the case for a non-heritable variance-covariance parameter $\Sigma_{e}$ if it is assumed to be the same for every observed species (i.e. tip) in the tree. 
- a local stacked parameter that has a different value for each regime in the model. For example, in an OU model, the selection strength matrix $H$ has a diffent value for each of the $R$ regimes in the model. Therefore, an OU model contains a $k\times k\times R$ array member called "H". The element `H[,,1]` would correspond to regime 1, `H[,,2]` to regime 2, etc. As a second example, the long-term optimum parameter of an OU process is a $k$-dimensional vector $\vec{\theta}_{r}$ for each regime in the model. In an OU model with $R$ regimes, this would be represented by a $k\times R$ array (matrix) called "theta", such that the vector `theta[, 1]` corresponds to regime 1. Finally, a local scalar parameter (i.e. a number), would be represented by an $R$-vector.
- a nested PCM object corresponding to a regime. This is the case for a mixed Gaussian phylogenetic model, where different types of Gaussian processes can be acting on different parts of the tree, represented by different regimes. 

So let us create

```{r}
modelObject <- structure(
  list(X0 = structure(c(0.0, 0.2),
                      class = c("VectorParameter", "_Global", "numeric")),
       Sigma = structure(abind(matrix(c(1,0,0,1), 2, 2),
                               matrix(c(2,0,0,2), 2, 2), along = 3),
                         class = c("MatrixParameter", "_UpperTriangularWithDiagonal", "_WithNonNegativeDiagonal", "matrix"))),
  class = c("PCM"),
  k = 2,
  regimes = 1:2)

PCMRegimes(modelObject)
PCMParamCount(modelObject)
PCMParamGetShortVector(modelObject)

vec <- double(PCMParamCount(modelObject))
PCMParamLoadOrStore(modelObject, vec, offset = 0, load=FALSE)
vec

modelObjectLowerLimit <- PCMParamLowerLimit(modelObject)
PCMParamLoadOrStore(modelObjectLowerLimit, vec, 0, load=FALSE)
vec

matParams <- PCMParamRandomVecParams(modelObject, n = 10)

matParamsX0 <- PCMParamRandomVecParams(modelObject$X0, 2, 2, n = 10)

matParamsSigma <- PCMParamRandomVecParams(modelObject$Sigma, 2, 2, n = 10)
```
## Specifying a model type
```{r}
library(data.table)
OUModelDummy <- list()
class(OUModelDummy) <- c("OU")

listParameterizationsOU <- PCMListParameterizations(OUModelDummy)
dtParameterizations <- PCMTableParameterizations(OUModelDummy)

PCMGenerateParameterizations(OUModelDummy, tableParameterizations = dtParameterizations[1:10])

attr(OUModelDummy, "k") <- 2
attr(OUModelDummy, "regimes") <- letters[1:3]
OUObject <- PCM("OU", k = 2, regimes = letters[1:3])


OUObject2 <- PCM("OU__Omitted_X0__H__Theta__UpperTriangularWithDiagonal_WithNonNegativeDiagonal_Sigma_x__UpperTriangularWithDiagonal_WithNonNegativeDiagonal_Sigmae_x", k = 2, regimes = letters[1:3])
```
```{r}
library(data.table)
OUModelDummy <- list()
class(OUModelDummy) <- c("OU")

listParameterizationsOU <- PCMListParameterizations(OUModelDummy)
dtParameterizations <- PCMTableParameterizations(OUModelDummy)

PCMGenerateParameterizations(OUModelDummy, tableParameterizations = dtParameterizations[1:10])

MGObject <- MixedGaussian(k = 2, modelTypes =  PCMModels(), mapping = c(1, 3, 5))

ul <- PCMParamUpperLimit(MGObject)
rv <- PCMParamRandomVecParams(MGObject)
PCMParamGetShortVector(ul)
```
