---
title: "A User Guide to the PCMBase R-package"
author: "Venelin Mitov"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{A User Guide to the PCMBase R-package}
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(phytools)
library(abind)
library(PCMBase)
library(POUMM)
```


# Data 

The input data for the phylogenetic comparative methods covered in the PCMBase package consists of a phylogenetic tree of $N$ tips and a $k\times N$ matrix, $X$ of observed trait-values, where $k$ is the number of traits. The matrix $X$ can contain ```NA```s corresponding to missing values.

# Models
A *model* is defined by a set of parameters and a rule stating how these parameters should be fit to the data. For example, a multivariate phylogenetic Ornstein-Uhlenbeck mixed model has the following parameters:

- ```X0``` : a $k$-vector of initial values;
- ```H``` : a $k\times k$ matrix denoting the selection strength of the process;
- ```Theta``` : a $k$-vector of long-term optimal trait values;
- ```Sigma``` : a $k\times k$ matrix denoting the stochastic drift variance-covariances;
- ```Sigmae``` : a $k\times k$ matrix denoting the non-genetic (non-heritable) variance covariance;

The rule defining how the above OU-parameters should be fit to the data is defined by the multivariate OU Gaussian distribution, as described in (insert reference);

## Model regimes
Model *regimes* are different instances of the model parameters, which are associated with different branches of the phylogenetic tree. 

## Examples 

In PCMBase, models are specified as S3 objects, i.e. ordinary R-lists with a class attribute. Below we show how to specify an OU model:

```{r}
# number of regimes
R <- 2
# number of traits
k <- 3

## Specifying a trivariate OU process for each regime
# First, specify the X0, H, Theta, Sigma and Sigmae2 parameters for each regime.
# Then we use the abind function to stack the parameters into arrays which's last
# dimension is the regime

# regimes

# in regime 'a' the three traits evolve according to three independent OU processes
a.X0 <- c(5, 2, 1)
a.H <- rbind(c(0, 0, 0),
             c(0, 2, 0),
             c(0, 0, 3))
a.Theta <- c(10, 6, 2)
a.Sigma <- rbind(c(1.6, 0, 0),
                 c(0, 2.4, 0),
                 c(0, 0, 2))
a.Sigmae2 <- rbind(c(0, 0, 0),
                   c(0, 0, 0),
                   c(0, 0, 0))

# in regime 'b' there is a correlation between the traits
b.X0 <- c(12, 4, 3)
b.H <- rbind(c(2, .1, .2),
             c(.1, .6, .2),
             c(.2, .2, .3))
b.Theta <- c(10, 6, 2)
b.Sigma <- rbind(c(1.6, .3, .3),
                 c(.3, 0.3, .4),
                 c(.3, .4, 2))
b.Sigmae2 <- rbind(c(.2, 0, 0),
                   c(0, .3, 0),
                   c(0, 0, .4))

H <- abind::abind(a.H, b.H, along=3, 
                      new.names=list(x=NULL, y=NULL, regime=c('a','b')))
Theta <- abind::abind(a.Theta, b.Theta, along=2, 
                      new.names=list(xy=NULL, regime=c('a', 'b')))
Sigma <- abind::abind(a.Sigma, b.Sigma, along=3, 
                      new.names=list(x=NULL, y=NULL, regime=c('a','b')))
Sigmae <- abind::abind(a.Sigmae2, b.Sigmae2, along=3, 
                       new.names=list(x=NULL, y=NULL, regime=c('a','b')))


## Simulations of trait data

# regime 'a', trait 1
model.a.1 <- list(X0 = a.X0[1], 
                  H=H[1,1,'a',drop=FALSE],
                  Theta=Theta[1,'a',drop=FALSE],
                  Sigma=Sigma[1,1,'a',drop=FALSE],
                  Sigmae=Sigmae[1,1,'a',drop=FALSE])
class(model.a.1) <- 'OU'

# regime 'a', trait 2
model.a.2 <- list(X0 = a.X0[2], 
                  H=H[2,2,'a',drop=FALSE],
                  Theta=Theta[2,'a',drop=FALSE],
                  Sigma=Sigma[2,2,'a',drop=FALSE],
                  Sigmae=Sigmae[2,2,'a',drop=FALSE])
class(model.a.2) <- 'OU'

# regime 'a', trait 3
model.a.3 <- list(X0 = a.X0[3], 
                  H=H[3,3,'a',drop=FALSE],
                  Theta=Theta[3,'a',drop=FALSE],
                  Sigma=Sigma[3,3,'a',drop=FALSE],
                  Sigmae=Sigmae[3,3,'a',drop=FALSE])
class(model.a.3) <- 'OU'

# regime 'a', traits 1, 2 and 3
model.a.123 <- list(X0 = a.X0,
                    H=H[,,'a',drop=FALSE],
                    Theta=Theta[,'a',drop=FALSE],
                    Sigma=Sigma[,,'a',drop=FALSE],
                    Sigmae=Sigmae[,,'a',drop=FALSE])
class(model.a.123) <- 'OU'


# regime 'b', traits 1, 2 and 3
model.b.123 <- list(X0 = b.X0,
                    H=H[,,'b',drop=FALSE],
                    Theta=Theta[,'b',drop=FALSE],
                    Sigma=Sigma[,,'b',drop=FALSE],
                    Sigmae=Sigmae[,,'b',drop=FALSE])
class(model.b.123) <- 'OU'

# regimes 'a' and 'b', traits 1, 2 and 3
model.ab.123 <- list(X0 = a.X0, 
                     H=H[,,,drop=FALSE],
                     Theta=Theta[,,drop=FALSE],
                     Sigma=Sigma[,,,drop=FALSE],
                     Sigmae=Sigmae[,,,drop=FALSE])
class(model.ab.123) <- 'OU'

```


# Simulating data

## Single regime simulations
```{r phylo}
set.seed(1)

library(phytools)

# number of tips
N <- 30

tree.a <- phytools::pbtree(n=N, scale=1)

# tree with one regime
tree.a <- phytools::pbtree(n=N, scale=1)

plot(tree.a)

# generate traits
traits.a.1 <- mvsim(tree.a, model.a.1, 0)
traits.a.2 <- mvsim(tree.a, model.a.2, 0)
traits.a.3 <- mvsim(tree.a, model.a.2, 0)
```

## Two regimes simulations
```{r}
# number of regimes
R <- 2

# rate mtrix of transition from one regime to another
Q <- matrix(c(-1, 1, 1, -1), R, R)
colnames(Q) <- rownames(Q) <- letters[1:R]
colnames(Q) <- rownames(Q) <- letters[1:R]

tree.ab <- phytools::sim.history(tree.a, Q, anc='a')

# convert the simmap tree to a normal phylo object with singleton nodes at the
# within-branch regime changes. The regimes are encoded as names of the 
# edge.length vector
tree.ab <- map.to.singleton(tree.ab)

plotTree.singletons(tree.ab)

traits.a.123 <- mvsim(tree.a, model.a.123, c(0,0,0))
```

# Calculating likelihoods
```{r}
# For the single trait models the likelihoods calculated with mvlik and 
# POUMM::likPOUMMGivenTreeVTips should match exactly:
cat("PCMBase: \n")
mvlik(traits.a.1$values + traits.a.1$errors, tree.a,
              model.a.1)
cat("POUMM: \n")
POUMM::likPOUMMGivenTreeVTips(
            z = traits.a.1$values[1,] + traits.a.1$errors[1,],
            tree = tree.a,
            alpha = model.a.1$H[1,1,1],
            theta = model.a.1$Theta[1,1],
            sigma = sqrt(model.a.1$Sigma[1,1,1]),
            sigmae = sqrt(model.a.1$Sigmae[1,1,1]),
            g0 = a.X0[1])

cat("PCMBase: \n")
mvlik(traits.a.2$values + traits.a.2$errors, tree.a,
              model.a.2)
cat("POUMM: \n")
POUMM::likPOUMMGivenTreeVTips(
            z = traits.a.2$values[1,] + traits.a.2$errors[1,],
            tree = tree.a,
            alpha = model.a.2$H[1,1,1],
            theta = model.a.2$Theta[1,1],
            sigma = sqrt(model.a.2$Sigma[1,1,1]),
            sigmae = sqrt(model.a.2$Sigmae[1,1,1]),
            g0 = a.X0[2])

cat("PCMBase: \n")
mvlik(traits.a.3$values + traits.a.3$errors, tree.a,
              model.a.3)
cat("POUMM: \n")
POUMM::likPOUMMGivenTreeVTips(
            z = traits.a.3$values[1,] + traits.a.3$errors[1,],
            tree = tree.a,
            alpha = model.a.3$H[1,1,1],
            theta = model.a.3$Theta[1,1],
            sigma = sqrt(model.a.3$Sigma[1,1,1]),
            sigmae = sqrt(model.a.3$Sigmae[1,1,1]),
            g0 = a.X0[3])

# For the multivariate model, the likelihood should be equal to the sum of the
# corresponding univariate likelihoods since the traits are independent:

cat("PCMBase: \n")
mvlik(X = traits.a.123$values + traits.a.123$errors, 
      tree = tree.a, 
      model = model.a.123)
cat("POUMM: \n")
POUMM::likPOUMMGivenTreeVTips(
            traits.a.123$values[1,]+traits.a.123$errors[1,],
            tree.a,
            model.a.123$H[1,1,1],
            model.a.123$Theta[1,1],
            sqrt(model.a.123$Sigma[1,1,1]),
            sqrt(model.a.123$Sigmae[1,1,1]),
            a.X0[1]) +
             POUMM::likPOUMMGivenTreeVTips(
               traits.a.123$values[2,] + traits.a.123$errors[2,],
               tree.a,
               model.a.123$H[2,2,1],
               model.a.123$Theta[2,1],
               sqrt(model.a.123$Sigma[2,2,1]),
               sqrt(model.a.123$Sigmae[2,2,1]),
               a.X0[2]) +
             POUMM::likPOUMMGivenTreeVTips(
               traits.a.123$values[3,]+traits.a.123$errors[3,],
               tree.a,
               model.a.123$H[3,3,1],
               model.a.123$Theta[3,1],
               sqrt(model.a.123$Sigma[3,3,1]),
               sqrt(model.a.123$Sigmae[3,3,1]),
               a.X0[3])

```
